Uncovering underlying musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and aural music skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks

### 2023-04-19 discussion/discussion.tex
Updates from Overleaf

\chapter{Results}

This is an example paragraph. As you can see, the main text uses a font size of 12 pt and a line spacing of 1.5. Neither the paragraphs nor the first lines of paragraphs should be indented.

There is no very strict page limit. Your number of pages will be strongly influenced by the size and total number of your figures and tables. It is recommended staying within 30-50 pages. Do not try to fill as many pages as you can. Longer theses are not necessarily of higher quality and of more non-redundant content than shorter theses. Certainly, a master thesis of 15 pages is too short, and a master thesis of 100 pages is too long.

\section{Tables and graphics}

This is an example paragraph. As you can see, the main text uses a font size of 12 pt and a line spacing of 1.5. Neither the paragraphs nor the first lines of paragraphs should be indented.

There is no very strict page limit. Your number of pages will be strongly influenced by the size and total number of your figures and tables. It is recommended staying within 30-50 pages. Do not try to fill as many pages as you can. Longer theses are not necessarily of higher quality and of more non-redundant content than shorter theses. Certainly, a master thesis of 15 pages is too short, and a master thesis of 100 pages is too long.

\begin{figure}[!ht]
\includegraphics[clip,width=\columnwidth]{Figures/PlotTimeSeriesResult}% 
\caption{This is an example of a figure and its caption.}
\label{fig:timeseries}
\end{figure}

\begin{table}[!ht]
\renewcommand{\arraystretch}{1.50}
\caption{This is an example of a table and its caption.}
\label{tablePCA}
\centering
\begin{tabular}{| c | c |}
\hline
\bfseries PCA & \bfseries Residual mean (in absolute values) \\
\hline\hline
Original PCA & 0.1267  \\
\hline
PCA on Centroid 1 & 0.1249\\
\hline
PCA on Centroid 2 & 0.1214  \\
\hline
\end{tabular}
\end{table}

>>>>

\chapter{Discussion}

XXXXXXXXXXX

\section{Discussion}

XXXXXXXXXXX

\section{Conclusions}

XXXXXXXXXXX

### 2023-04-19 introduction/Subsections/Motivation.tex
Updates from Overleaf

\section{About music aural skills and high-level perceptual concepts}

@@

Schenkerian Analysis, devised by Austrian music theorist Heinrich Schenker, explores tonal music's underlying structure by concentrating on the hierarchical relationships between pitches. Based on the idea that all tonal music shares a fundamental structure called "Ursatz" or "basic shape," the method identifies the stepwise descending line (Urlinie) and bass arpeggiation (Bassbrechung). Schenkerian Analysis simplifies a musical piece to its core elements, allowing analysts to examine how different layers contribute to the overall structure. Key concepts include prolongational levels, harmony, counterpoint, voice-leading, and Ursatz (fundamental structure). Although criticized for reducing all tonal works to a few background structures, Schenker emphasizes the importance of individual elaboration in determining a work's identity and meaning.

Anstieg (Initial Ascent): Ascending motion leading to the primary tone of the fundamental line.
Arpeggiation (Brechung): Elementary elaboration of a harmony.
Background (Hintergrund): The structural level of the fundamental structure.
Bass Arpeggiation (Bassbrechung): Bass pattern I-V-I forms the harmonic content of the background of tonal musical pieces.
Chord of Nature (Klang): The complex sound of the first five harmonic series notes suggests a model for the major triad.
Diminution: The process by which an interval formed by notes of longer value is expressed in notes of smaller value.
Foreground (Vordergrund): The surface level of musical structure in Schenkerian analysis.
Fundamental Line (Urlinie): The melodic aspect of the fundamental structure, a stepwise descent from one of the triad notes to the tonic.
Middleground (Mittelgrund): An intermediate structural level between the background and foreground in Schenkerian analysis.
Prolongation (Auskomponierung): The process in tonal music through which a pitch, interval, or consonant triad can govern music spans when not physically sounding.
Structural Level (Schicht): Refers to the different levels of musical structure in Schenkerian analysis, including background, middle-ground, and foreground.
Tonal Space: A general principle of Schenkerian analysis where intervals between the notes of the tonic triad form a tonal space filled with passing and neighbor notes.
Unfolding (Ausfaltung): The transformation of a single chord into a horizontal succession.
Voice Leading: The study of the principles that govern the progression of the component voices of the composition both separately and in combination.
The glossary also includes related articles, references, and external links for further reading and understanding Schenkerian analysis.

%%%%%%%%%%%%%
%MESOSTRUC
%%%%%%%%%%%%%

Mesostructures \cite{Mesostructures2023} are intermediate levels of articulation between waveshapes' microstructure and musical forms' macrostructure. Examples of mesostructures include melody, arpeggios, syncopation, polyphonic grouping, and textural contrast. The article notes that mesostructures have received limited attention in deep learning, as current models only focus on microstructures. The article proposes a new approach to mesostructural audio modeling, which involves a differentiable arpeggiator and time-frequency scattering. The report also discusses the need for a time-invariant and multiscale differentiable time-frequency model of similarity at the level of both local spectra and spectrotemporal modulations.

@@

\subsection{Sample MIR task}

>>>>

\subsection{About music aural skills and high-level perceptual concepts}

@@

Schenkerian Analysis, developed by Austrian music theorist Heinrich Schenker, investigates the underlying structure of tonal music by focusing on the hierarchical relationships between pitches. This analytical approach is founded on the premise that all tonal music shares a fundamental structure known as "Ursatz" or "basic shape." The method identifies a piece's stepwise descending line (Urlinie) and bass arpeggiation (Bassbrechung), simplifying it to its essential elements and examining how different layers contribute to the overall structure. Central concepts in Schenkerian Analysis include prolongation levels, harmony, counterpoint, voice-leading, and Ursatz (fundamental structure). Despite criticism for reducing all tonal works to a limited number of background structures, Schenker highlights the significance of individual elaboration in determining a piece's identity and meaning.

While such analysis is designed for the symbolic domain, we argue that musical abstraction is also feasible in the audio time domain. In this section, we explore the application of Schenkerian Analysis principles to the Music Information Retrieval (MIR) domain by examining the foreground, middle-ground, and background levels of musical structure as follows:

\begin{itemize}
\item \textbf{Foreground}: The surface level of a musical composition, consisting of the notes, rhythms, and articulations directly perceived by the listener, representing the most detailed level of the musical structure.

\item \textbf{Middle-ground and mesostructures}: An intermediate level of analysis focusing on essential structural elements such as harmonic progressions, thematic materials, and voice-leading reductions that contribute to the overall composition. In the MIR domain, this level can also be understood as mesostructure. Mesostructures are intermediate levels of articulation between waveshapes' microstructure and musical forms' macrostructure. Examples of mesostructures include melody, arpeggios, syncopation, polyphonic grouping, and textural contrast. Mesostructures might have received limited attention in deep learning, as current models only focus on microstructures \cite{Mesostructures2023}.

\item \textbf{Background}: The most abstract level of musical structure, representing the fundamental core consisting of the \textit{Urlinie} and \textit{Bassbrechung}, which serves as the foundation for the entire composition.
\end{itemize}

By understanding and applying Schenkerian Analysis principles to the MIR domain, we can better understand the complex relationships between different levels of musical structure in the audio time domain, allowing for a more comprehensive analysis of tonal music.

@@

\subsubsection{Sample MIR task}

### 2023-04-19 introduction/Subsections/Objectives.tex
Updates from Overleaf

In recent years, using neural networks has enabled low-dimensional musical embeddings that capture crucial musical features and allow the computation of content-based similarity between musical passages \cite{}. By acquiring an embedding function $f(x) \in \mathbb{R}^d$ that maps a given course $x$ to a vector in the real space $\mathbb{R}^d$, we can measure the similarity between two musical passages using the Euclidean distance, as represented by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The effectiveness of such systems depends on the level of musical abstraction and the embedding space's ability to interpret music in a way that aligns with human perception. Therefore, a suitable embedding space is crucial to ensure the output aligns with human expectations.

@@

\begin{enumerate}
\item Music recommendation: The learned embedding space can retrieve similar musical content and recommend songs or playlists to users based on their listening history or preferences.
\item Music classification: The embedding space can classify songs based on genre, mood, or other musical characteristics.
\item Music transcription: The learned embedding space can transcribe audio signals into symbolic music notation, such as MIDI or sheet music.
\item Music generation: The embedding space can generate new musical content by sampling from the learned distribution of musical features.
\item Music synchronization: The discovered embedding space can synchronize audio and visual content in applications such as music videos or virtual reality experiences.
\item Music search: The embedding space can search for specific musical content based on high-level perceptual concepts and the relationships they have among them
\end{enumerate}

>>>>

In recent years, using neural networks has enabled low-dimensional musical embeddings that capture crucial musical features and allow the computation of content-based similarity between musical passages \cite{Chen2020LearningRecommendation}\cite{KimLEARNINGLOSS}\cite{Hung2022Feature-informedClassification}\cite{WonEmotionStories}. By acquiring an embedding function $f(x) \in \mathbb{R}^d$ that maps a given course $x$ to a vector in the real space $\mathbb{R}^d$, we can measure the similarity between two musical passages using the Euclidean distance, as represented by the formula $D_{i,j} = ||f(x_i) - f(x_j)||^2$. The effectiveness of such systems depends on the level of musical abstraction and the embedding space's ability to interpret music in a way that aligns with human perception. Therefore, a suitable embedding space is crucial to ensure the output aligns with human expectations.

@@

### 2023-04-19 methods/subsections/experiment setup and hyperparameter tuning.tex
Updates from Overleaf

For finetuning and improving the model's behavior, we followed \href{http://karpathy.github.io/2019/04/25/recipe/}{\textit{A Recipe for Training Neural Networks}} \cite{AndrejKarpathyHttp://karpathy.github.io/2019/04/25/recipe/} by \href{https://karpathy.ai/}{Andrej Karpathy} \cite{AndrejKarpathyHttps://karpathy.ai/}. A blog post that discusses a recipe for training neural networks. The author of the blog aims to expand on common mistakes people make when training neural nets by explaining how to avoid them through a particular process. Andrej Karpathy notes that there is often a gap between understanding the technical aspects of a neural network and achieving state-of-the-art results.

\input{figures/neural networks/dropout}

>>>>

Musically motivated NN \cite{musicmotivCNN}.

For finetuning and improving the model's behavior, we followed \href{http://karpathy.github.io/2019/04/25/recipe/}{\textit{A Recipe for Training Neural Networks}} \cite{AndrejKarpathyHttp://karpathy.github.io/2019/04/25/recipe/} by Andrej Karpathy. A blog post that discusses a recipe for training neural networks. The author of the blog aims to expand on common mistakes people make when training neural nets by explaining how to avoid them through a particular process. Andrej Karpathy notes that there is often a gap between understanding the technical aspects of a neural network and achieving state-of-the-art results.

### 2023-04-19 methods/subsections/machine learning.tex
Updates from Overleaf

\section{Introduction}
Machine Learning (ML), particularly Deep Learning, has significantly advanced the MIR field by enabling automatic feature learning, improving classification and analysis tasks, and facilitating novel applications like music generation and source separation. Adopting musically motivated neural network architectures \cite{musicmotivCNN} blends musical concepts and perceptual approaches and use this knowledge to analyze and understand music in a way that goes beyond traditional music analysis techniques.

>>>>

\section{Introduction}

In this chapter, we present the methodology employed in this study.

We provide an overview of the Machine Learning (ML) and Deep Learning techniques that have played a pivotal role in advancing the MIR field. We then discuss the importance of feature extraction and selection in MIR and how automatic feature learning has facilitated the development of more robust models.

Subsequently, we delve into the concept of musically motivated neural network architectures, which incorporate domain-specific knowledge and musical ideas into the design of neural networks. We explore techniques proposed in the literature for blending musical concepts and perceptual approaches.

We then outline the specific MIR tasks that our study focuses on and discuss the evaluation metrics used to assess the performance of the proposed models. Finally, we present the experimental setup, including the datasets used for training and validation, the preprocessing steps, and the training parameters.

Throughout this chapter, we aim to provide a comprehensive understanding of the methodology employed in our research, laying the foundation for the presentation and discussion of the results in the following chapters.

### 2023-04-19 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

The SampleCNN model takes in 1D input data with a single channel. The first layer is a 1D convolutional layer with a kernel size of 3, stride of 3, and 128 output channels. It is followed by batch normalization and ReLU activation.

After the initial layer, there are 9 hidden layers with varying kernel sizes, strides, and channels according to the values specified in self.strides and self.hidden. Each hidden layer consists of:

A 1D convolutional layer with the specified number of input and output channels, kernel size, stride of 1, and padding of 1.
Batch normalization.
ReLU activation.
Max pooling with kernel size and stride equal to the stride value of the current layer.
After the hidden layers, there's another 1D convolutional layer with 512 input channels, 512 output channels, kernel size of 3, stride of 1, and padding of 1. This is followed by batch normalization and ReLU activation.

@@

If the model is in supervised mode, dropout with a rate of 0.5 is applied.

@@

Layer 1:

Input channels: 1
Output channels: 128
Layer 2:

Input channels: 128
Output channels: 128
Layer 3:

Input channels: 128
Output channels: 128
Layer 4:

Input channels: 128
Output channels: 128
Layer 5:

Input channels: 128
Output channels: 256
Layer 6:

Input channels: 256
Output channels: 256
Layer 7:

Input channels: 256
Output channels: 256
Layer 8:

Input channels: 256
Output channels: 256
Layer 9:

Input channels: 256
Output channels: 256
Layer 10:

Input channels: 256
Output channels: 256
Layer 11:

Input channels: 256
Output channels: 256
Layer 12:

Input channels: 256
Output channels: 512
Layer 13:

Input channels: 512
Output channels: 512
Layer 14:

Input channels: 512
Output channels: 128

\subsection{Audio augmentation and transformation pipeline}
The positive samples are created by transforming and augmenting the original samples while preserving their perceptual underlying musical language. This is achieved using techniques such as time-stretching, pitch-shifting, and adding noise to the samples. By doing so, we aim to enhance the diversity of the positive samples and improve the model's ability to retrieve similar-sounding samples that share the same underlying musical language.

On the other hand, the negative samples are created by shuffling and scrambling the original samples while maintaining their production texture. This is achieved using random segment selection, time-domain filtering, and frequency-domain masking. The goal is to retain the texture production of the samples while disrupting their underlying composition.

@@

Implements the triplet loss function, commonly used for learning embeddings in metric learning problems. The triplet loss aims to ensure that the distance between an anchor and a positive sample (both from the same class) is smaller than the distance between the anchor and a negative sample (from a different class) by a margin.

>>>>

The SampleCNN model takes in 1D input data with a single channel. The first layer is a 1D convolutional layer with a kernel size of 3, a stride of 3, and 128 output channels. It is followed by batch normalization and ReLU activation.

After the initial layer, there are nine hidden layers with varying kernel sizes, strides, and channels. Each hidden layer consists of:

\begin{itemize}
    \item A 1D convolutional layer with the specified number of input and output channels, kernel size, a stride of 1, and padding of 1.
    \item Batch normalization.
    \item ReLU activation.
    \item Max pooling with kernel size and stride equal to the stride value of the current layer.
\end{itemize}

\begin{tabular}{|c|c|c|}
\hline
\textbf{Layer} & \textbf{Input Channels} & \textbf{Output Channels} \\
\hline
1 & 128 & 128 \\
\hline
2 & 128 & 128 \\
\hline
3 & 128 & 256 \\
\hline
4 & 256 & 256 \\
\hline
5 & 256 & 256 \\
\hline
6 & 256 & 256 \\
\hline
7 & 256 & 256 \\
\hline
8 & 256 & 256 \\
\hline
9 & 256 & 512 \\
\hline
\end{tabular}

After the hidden layers, there's another 1D convolutional layer with 512 input channels, 512 output channels, a kernel size of 3, a stride of 1, and padding of 1. This is followed by batch normalization and ReLU activation.

@@

If the model is "supervised", dropout with a rate of 0.5 is applied.

\input{figures/neural networks/dropout}

@@

\subsection{Audio augmentation and transformation pipeline}
\subsubsection{Positive sample generation}
Let $W$ be the input waveform and $W_p$ be the generated positive waveform. The pipeline applies a series of audio effects and adds noise with a random SNR. Define the following random variables for the effects:

\begin{alignat*}{2}
&gain \sim \text{Uniform}(-12, 0), \quad
&pitch \sim \text{Uniform}(-1200, 1200), \\
&reverb\_params \sim \text{Uniform}(0, 100)^3, \quad
&chorus\_params = \begin{cases}
\text{Uniform}(0.1, 1.0), & i=1,2, \\
\text{Uniform}(20, 55), & i=3, \\
\text{Uniform}(0.1, 0.9), & i=4, \\
\text{Uniform}(0.1, 2.0), & i=5, \\
\text{Uniform}(2, 5), & i=6, \\
\text{Categorical}(\{-s, -t\}), & i=7,
\end{cases}, \\
&drive \sim \text{Uniform}(0, 30), \quad
&stretch \sim \text{Uniform}(0.9, 1.1), \\
&speed \sim \text{Uniform}(0.9, 1.1), \quad
&tremolo\_speed \sim \text{Uniform}(0.1, 100), \\
&tremolo\_depth \sim \text{Uniform}(1, 101), \quad
&snr\_range \sim \text{Uniform}(12, 100).
\end{alignat*}

For white noise, let $N$ be the noise matrix with dimensions (channels, samples) as $W$. Calculate signal power $P_s$ and noise power $P_n$:

\begin{align*}
&P_s = \sum (W_{ij})^2, \
&P_n = \sum (N_{ij})^2.
\end{align*}

Calculate the scaling factor $\alpha$ for noise based on the desired SNR:

\[
\alpha = \sqrt{\frac{P_s}{P_n \times 10^{\frac{snr\_range}{10}}}}
\]

Compute the scaled noise matrix $N_s$ and the noisy waveform matrix $W_n$:

\begin{align*}
&N_s = \alpha \times N, \
&W_n = W + N_s.
\end{align*}

The final positive waveform $W_p$ is obtained with noise added according to the chosen SNR. This pipeline enhances the diversity of the positive samples and aims to improve the model's ability to retrieve similar-sounding samples sharing the same underlying musical language.

\subsubsection{Negative sample generation}

Let $a$ be the original anchor clip and $n$ be the negative clip to be generated. The steps for generating the negative clip $n$ are as follows:

\begin{enumerate}
\item Calculate the length $L_a$ and duration $T_a$ of the anchor clip $a$:

$L_a = \text{shape}(a)$, $\quad T_a = \frac{L_a}{f_s}$,

where $\text{shape}(a)$ returns the shape of the audio clip, and $f_s$ is the sample rate of the audio.

\item Determine the number of chunks $n_c$ to split the anchor clip into based on a minimum chunk duration $T_c$:
$$
n_c = \left\lfloor\frac{T_a}{T_c}\right\rfloor.
$$

\item Calculate the minimum and maximum chunk lengths in samples:
$$
L_{c,\text{min}} = T_c \times f_s, \quad L_{c,\text{max}} = T_{\text{max}} \times f_s,
$$
where $T_{\text{max}}$ is the maximum chunk duration specified in seconds.

\item Generate $n_c$ chunk lengths $L_{c,1},\ldots,L_{c,n_c}$ as follows:
$$
L_{c,i} = \begin{cases}
L_{c,\text{min}}, & i=1, \\
\text{randint}(L_{c,\text{min}}, L_{c,\text{max}}), & 1<i<n_c, \\
L_a - \sum_{j=1}^{i-1} L_{c,j}, & i=n_c.
\end{cases}
$$

\item Split the anchor clip into chunks $c_1,\ldots,c_{n_c}$ based on the lengths $L_{c,1},\ldots,L_{c,n_c}$:
$$
c_i = a[..., s_i : s_i + L_{c,i}], \quad S_c = \text{cumsum}([0] + [L_{c,1},\ldots,L_{c,n_c-1}]), \quad s_i = S_c[i], \quad i=1,\ldots,n_c.
$$

\item Shuffle the resulting list of chunks randomly to obtain the reordered chunks $c'_1,\ldots,c'_{n_c}$:
$$
c' = \text{shuffle}([c_1,\ldots,c_{n_c}]).
$$

\item Concatenate the shuffled chunks $c'_1,\ldots,c'_{n_c}$ to create the negative clip $n$:
$$
n = \text{concatenate}(c').
$$

\end{enumerate}

@@

The triplet loss aims to ensure that the distance between an anchor and a positive sample (both from the same class) is smaller than the distance between the anchor and a negative sample (from a different class) by a margin.
